---
title: "MB5370 Module 1: Intro to programming"
author: "Katie Dyck"
output: html_document
date: "2026-05-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
# load packages

library(tidyverse)
library(palmerpenguins)
```

```{r}
getwd()
```

## WORKSHOP 01. INTRO

This section contains code and programming from the first workshop. A basic introduction to R.

```{r Workshop 1 Code}
## ------ Getting Started ------ ####
# This section introduces us to R by using it as a simple calculator.

2 + 1
1:30
# 6 *
  # not complete - when you run this line, console asks you to unput a number after the asterisk

# 6 % 4  - doesn't work!
6 / 4



## ------ Variables and Assignment ------ ####
# This section focuses on understanding how data is stored in R and why that matters.

age <- 25
first_name <- 'Bill'

age
first_name

age + 1
age + age

# In R, you can’t use a number at the beginning of an object name and you can’t use some special symbols, including spaces.
# 01_age <- 25 # starts with a number
# !_age <- 25 # no special symbols
# age bob <- 25 # no spaces
Age <- 41 # cases matter

`age bob` <- 25 # no spaces, but with back ticks
# This is useful in some situations, however, it is better to use informative object names to make your code readable

# Run an algorithm:
25 + 15.1 + 20.25
x <- 30
25 + 15.1 + 20.25 + x



## ------ Functions ------ ####
# This section focuses on exploring built-in R functions

years_old <- 25.7
round(years_old)  # round up
floor(years_old)  # returns the largest integer not greater than elements of 'years_old'

years_old <- 25.765
round(years_old, digits = 2)

age <- 30
name <- "Katie"
paste(name, "is", age, "years old.", sep = " ")
paste(name, "is almost", age + 1, "years old.", sep = " ")



## ------ Debugging ------ ####
# Debugging is the process of finding and fixing errors in your code.

# Misconceptions
# variables don't work the same way as they do in spreadsheets

grade <- 55
total <- grade + 10
print(total)
grade <- 90
print(total)  # value of total in a spreadsheet would be 100, but here a variable holds the value it was assigned (65)
total <- grade + 10
print(total)

p <- 2
z <- 5
out <- p * z
print(out)

# debugging errors
# out <- p * a   # undefined variable 'a'
## > Error: object 'a' not found
out <- p * z

x <- 1
is.character(x)
is.integer(x)
is.numeric(x)
is.logical(x)

# my_quiz <- c("uno",
#              "dos",
#              "tres",
#              "cuatro"   # comma is missing!
#              "cinco")
# print(my_quis)   # spelling error
# str(my_quiz)
# len(my_quiz)   # len() not a function

my_quiz_fixed <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print(my_quiz_fixed)
str(my_quiz_fixed)
length(my_quiz_fixed)



## ------ Data types ------ ####

my_numeric_variable <- 4.2
class(my_numeric_variable)
typeof(my_numeric_variable)
attributes(my_numeric_variable)
length(my_numeric_variable)



## ------ Data structures ------ ####

# A vector is a collection of elements that are most commonly of the type character, numeric, integer, logical.
y <- c(1,2,3)
class(y)
typeof(y)
z <- c("Sarah", "Tracy", "John")
class(z)
typeof(z)

# Values inside a list can be several different types.
x <- list(1, "A", TRUE)
x
class(x)
typeof(x)
length(x)
x[[2]]   # use square brackets to retrieve individual elements in a list



## ------ Data frames and tibbles ------ ####
# 2D or "rectangular" data files

my_dataframe <- data.frame(
  no = c(1,2,3),
  name = c("Tracey", "John", "Pete"),
  logical = c(TRUE, FALSE, TRUE)
)
my_dataframe
str(my_dataframe)

my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)



## ------ Packages and libraries ------ ####
# Using software that other people have written

# install.packages("ggplot2", "tidyr", "earthtones", "dplyr")
# library(ggplot2)



## ------ Coding best practices ------ ####

# To take a .csv file and turn it into a plot of the abundance of a single species:
#   1. read the .csv file dataset into R - read_csv()
#   2. view the dataset - e.g. str('dataset_name_here'), head('dataset_name_here'), glimpse('dataset_name_here')
#   3. identify which columns/variables you want to use to create the plot, and which plot type (geom_*()) you want to create
#   4. create your plot with ggplot() + geom_*()
```


## WORKSHOP 02. VISUALISATION

```{r mpg dataset}
str(mpg)
# "displ" = engine size (litres)
# "hwy" = car's fuel efficiency (mpg)
```

### - Create a ggplot -

```{r Basic ggplot's}
# These create the exact same plot:
ggplot(mpg) +
    geom_point(aes(x = displ, y = hwy))
ggplot(mpg, aes(displ, hwy)) +
    geom_point()

# change transparency by class (alpha)
ggplot(mpg, aes(displ, hwy, alpha = drv)) +
    geom_point()

# change colour by class
ggplot(mpg, aes(displ, hwy, color = drv)) +
    geom_point()

# change point shape by class
ggplot(mpg, aes(displ, hwy, shape = drv)) +
    geom_point()

# change point size by class
# change colour by class
ggplot(mpg, aes(displ, hwy, size = drv)) +
    geom_point()

# change shape, size, AND colour by class
ggplot(mpg, aes(displ, hwy, color = drv, shape = drv, size = drv)) +
    geom_point()

ggplot(mpg, aes(displ, hwy, color = drv, shape = drv, size = drv, alpha = drv)) +
    geom_point()
# adding "alpha" isn't necessary in this case

ggplot(mpg, aes(displ, hwy, colour = displ < 5)) +
    geom_point()
```

```{r facet_wrap() and facet_grid()}
ggplot(mpg, aes(displ, hwy, colour = drv)) +
    geom_jitter() +
    facet_wrap(~ drv) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold")) +
    labs(title = "Fuel efficiency vs. engine size, sorted by drive train type",
         x = "Engine displacement (L)",
         y = "Highway (mpg)")

ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_wrap(~ class, nrow = 2)

# facet_grid() is used with more than one variable
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(drv ~ cyl)

# use a . if you don't want to facet in the rows or columns dimension
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(. ~ cyl)
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(drv ~ .)
```

```{r}
ggplot(mpg) +
    geom_smooth(aes(displ, hwy))

ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE)  # lm = 
```

```{r}
ggplot(mpg) +
    geom_smooth(aes(displ, hwy, linetype = drv))
```

You can create plots with multiple geoms to show relationships between different variables in the data:

```{r Using multiple geoms}
ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = drv)) +
    geom_smooth(aes(displ, hwy, colour = drv, linetype = drv))
```

You can use mapping in specific layers to enable the display of different aesthetics in different layers:

```{r Advanced plots}
ggplot(mpg, aes(displ, hwy)) +
    geom_smooth() +
    geom_point(aes(colour = class))

ggplot(mpg, aes(displ, hwy)) +
    geom_point(aes(colour = class)) +
    geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) +
    labs(caption = "Data has been filtered for class 'subcompact'")
```

### - Transformations and Stats -

```{r diamonds dataset}
str(diamonds)
```

```{r 'diamonds' plot 1}
ggplot(diamonds) +
    geom_bar(aes(x = cut))
```

-   *bar charts*, *histograms*, and *frequency polygrams* "bin" data and then plot bin counts (the number of data points that fall in each bin).
-   *smoothers* fit a model to the data and then plot predictions from the model.
-   *boxplots* compute a summary of the distribution and then display a specially formatted box.

`geoms` and `stats` can generally be used interchangeably.

```{r 'diamonds' plot 2}
ggplot(diamonds) +
    stat_count(mapping = aes(x = cut))
```

```{r Overriding ggplot defaults}
demo <- tribble(
    ~cut, ~freq,
    "Fair", 1610,
    "Good", 4906,
    "Very Good", 12082,
    "Premium", 13791,
    "Ideal", 21551
)
demo

ggplot(demo) +
    geom_bar(aes(x = cut, y = freq), stat = "identity")

# override default mapping from transformed variables to aesthetics
# e.g. display a bar chart of the PROPORTION of your total diamond dataset rather than a count
ggplot(diamonds) +
    geom_bar(aes(x = cut, y = stat(prop), group = 1))

# you can add statistical details to plots:
ggplot(diamonds) +
    stat_summary(
        mapping = aes(x = cut, y = depth),
        fun.min = min,
        fun.max = max,
        fun = median
    )
```

```{r Aesthetic adjustments}
# for bar plots, 'colour' changes the OUTLINE colour
ggplot(diamonds) +
    geom_bar(mapping = aes(x = cut, colour = cut))

# 'fill' changes the FILL colour
ggplot(diamonds) +
    geom_bar(mapping = aes(x = cut, fill = cut))

# default: position = "identity"
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity)) 

# alter transparency
ggplot(diamonds, aes(x = cut, fill = clarity)) +
    geom_bar(alpha = 1/5, position = "identity")

# colour bar outlines with no fill colour
ggplot(diamonds, aes(x = cut, colour = clarity)) +
    geom_bar(fill = NA, position = "identity")

# position = "fill" makes each set of stacked bars the same height
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "fill")

ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "dodge")

# position = "jitter" adds small amount of random noise to each point to avoid overplotting when points overlap...good for scatterplots, not barplots
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "jitter")
```
```{r 'penguins' dataset}
str(penguins)

ggplot(penguins, aes(x = species, fill = species)) +
    geom_bar()

ggplot(penguins, aes(x = species, y = sex, fill = island)) +
    geom_col(position = "dodge")

ggplot(penguins, aes(x = species, y = flipper_length_mm, colour = species)) +
    geom_jitter() +
    theme(legend.position = "none")
```


## WORKSHOP 03. REPRODUCIBLE RESEARCH IN MARINE SCIENCE

```{r linking GitHub and RStudio}
library(usethis)

# in the console:
  # credentials::git_credential_ask()
  # usethis::git_sitrep()

# in the console:
  # usethis::usegit()
```