---
title: "Module 1: Introduction to Programming"
author: "Katie Dyck"
date: "2026-05-20"
oexecute:
  echo: false
format:
    html:
        code-fold: show
        code-summary: "Show code"
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        number-sections: true
        number-depth: 4
        toc-location: left
        toc-expand: 2
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
# load packages

library(tidyverse)
library(palmerpenguins)
library(purrr)
```

```{r}
getwd()
```

# **Workshop 1: Introduction**

This section contains code and programming from the first workshop. A basic introduction to R.

```{r Workshop 1 Code}
## ------ Getting Started ------ ####
# This section introduces us to R by using it as a simple calculator.

2 + 1
1:30
# 6 *
  # not complete - when you run this line, console asks you to unput a number after the asterisk

# 6 % 4  - doesn't work!
6 / 4



## ------ Variables and Assignment ------ ####
# This section focuses on understanding how data is stored in R and why that matters.

age <- 25
first_name <- 'Bill'

age
first_name

age + 1
age + age

# In R, you can’t use a number at the beginning of an object name and you can’t use some special symbols, including spaces.
# 01_age <- 25 # starts with a number
# !_age <- 25 # no special symbols
# age bob <- 25 # no spaces
Age <- 41 # cases matter

`age bob` <- 25 # no spaces, but with back ticks
# This is useful in some situations, however, it is better to use informative object names to make your code readable

# Run an algorithm:
25 + 15.1 + 20.25
x <- 30
25 + 15.1 + 20.25 + x



## ------ Functions ------ ####
# This section focuses on exploring built-in R functions

years_old <- 25.7
round(years_old)  # round up
floor(years_old)  # returns the largest integer not greater than elements of 'years_old'

years_old <- 25.765
round(years_old, digits = 2)

age <- 30
name <- "Katie"
paste(name, "is", age, "years old.", sep = " ")
paste(name, "is almost", age + 1, "years old.", sep = " ")



## ------ Debugging ------ ####
# Debugging is the process of finding and fixing errors in your code.

# Misconceptions
# variables don't work the same way as they do in spreadsheets

grade <- 55
total <- grade + 10
print(total)
grade <- 90
print(total)  # value of total in a spreadsheet would be 100, but here a variable holds the value it was assigned (65)
total <- grade + 10
print(total)

p <- 2
z <- 5
out <- p * z
print(out)

# debugging errors
# out <- p * a   # undefined variable 'a'
## > Error: object 'a' not found
out <- p * z

x <- 1
is.character(x)
is.integer(x)
is.numeric(x)
is.logical(x)

# my_quiz <- c("uno",
#              "dos",
#              "tres",
#              "cuatro"   # comma is missing!
#              "cinco")
# print(my_quis)   # spelling error
# str(my_quiz)
# len(my_quiz)   # len() not a function

my_quiz_fixed <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print(my_quiz_fixed)
str(my_quiz_fixed)
length(my_quiz_fixed)



## ------ Data types ------ ####

my_numeric_variable <- 4.2
class(my_numeric_variable)
typeof(my_numeric_variable)
attributes(my_numeric_variable)
length(my_numeric_variable)



## ------ Data structures ------ ####

# A vector is a collection of elements that are most commonly of the type character, numeric, integer, logical.
y <- c(1,2,3)
class(y)
typeof(y)
z <- c("Sarah", "Tracy", "John")
class(z)
typeof(z)

# Values inside a list can be several different types.
x <- list(1, "A", TRUE)
x
class(x)
typeof(x)
length(x)
x[[2]]   # use square brackets to retrieve individual elements in a list



## ------ Data frames and tibbles ------ ####
# 2D or "rectangular" data files

my_dataframe <- data.frame(
  no = c(1,2,3),
  name = c("Tracey", "John", "Pete"),
  logical = c(TRUE, FALSE, TRUE)
)
my_dataframe
str(my_dataframe)

my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)



## ------ Packages and libraries ------ ####
# Using software that other people have written

# install.packages("ggplot2", "tidyr", "earthtones", "dplyr")
# library(ggplot2)



## ------ Coding best practices ------ ####

# To take a .csv file and turn it into a plot of the abundance of a single species:
#   1. read the .csv file dataset into R - read_csv()
#   2. view the dataset - e.g. str('dataset_name_here'), head('dataset_name_here'), glimpse('dataset_name_here')
#   3. identify which columns/variables you want to use to create the plot, and which plot type (geom_*()) you want to create
#   4. create your plot with ggplot() + geom_*()
```


# **Workshop 2: Data Visualization**

```{r mpg dataset}
str(mpg)
# "displ" = engine size (litres)
# "hwy" = car's fuel efficiency (mpg)
```

## Create a `ggplot`

```{r Basic ggplots}
# These create the exact same plot:
ggplot(mpg) +
    geom_point(aes(x = displ, y = hwy))
ggplot(mpg, aes(displ, hwy)) +
    geom_point()

# change transparency by class (alpha)
ggplot(mpg, aes(displ, hwy, alpha = drv)) +
    geom_point()

# change colour by class
ggplot(mpg, aes(displ, hwy, color = drv)) +
    geom_point()

# change point shape by class
ggplot(mpg, aes(displ, hwy, shape = drv)) +
    geom_point()

# change point size by class
# change colour by class
ggplot(mpg, aes(displ, hwy, size = drv)) +
    geom_point()

# change shape, size, AND colour by class
ggplot(mpg, aes(displ, hwy, color = drv, shape = drv, size = drv)) +
    geom_point()

ggplot(mpg, aes(displ, hwy, color = drv, shape = drv, size = drv, alpha = drv)) +
    geom_point()
# adding "alpha" isn't necessary in this case

ggplot(mpg, aes(displ, hwy, colour = displ < 5)) +
    geom_point()
```

```{r facet_wrap() and facet_grid()}
ggplot(mpg, aes(displ, hwy, colour = drv)) +
    geom_jitter() +
    facet_wrap(~ drv) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold")) +
    labs(title = "Fuel efficiency vs. engine size, sorted by drive train type",
         x = "Engine displacement (L)",
         y = "Highway (mpg)")

ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_wrap(~ class, nrow = 2)

# facet_grid() is used with more than one variable
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(drv ~ cyl)

# use a . if you don't want to facet in the rows or columns dimension
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(. ~ cyl)
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(drv ~ .)
```

```{r}
ggplot(mpg) +
    geom_smooth(aes(displ, hwy))

ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE)  # lm = 
```

```{r}
ggplot(mpg) +
    geom_smooth(aes(displ, hwy, linetype = drv))
```

You can create plots with multiple geoms to show relationships between different variables in the data:

```{r Using multiple geoms}
ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = drv)) +
    geom_smooth(aes(displ, hwy, colour = drv, linetype = drv))
```

You can use mapping in specific layers to enable the display of different aesthetics in different layers:

```{r Advanced plots}
ggplot(mpg, aes(displ, hwy)) +
    geom_smooth() +
    geom_point(aes(colour = class))

ggplot(mpg, aes(displ, hwy)) +
    geom_point(aes(colour = class)) +
    geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) +
    labs(caption = "Data has been filtered for class 'subcompact'")
```

## Transformations and Stats

```{r diamonds dataset}
str(diamonds)
```

```{r diamonds plot 1}
ggplot(diamonds) +
    geom_bar(aes(x = cut))
```

-   *bar charts*, *histograms*, and *frequency polygrams* "bin" data and then plot bin counts (the number of data points that fall in each bin).
-   *smoothers* fit a model to the data and then plot predictions from the model.
-   *boxplots* compute a summary of the distribution and then display a specially formatted box.

`geoms` and `stats` can generally be used interchangeably.

```{r diamonds plot 2}
ggplot(diamonds) +
    stat_count(mapping = aes(x = cut))
```

```{r Overriding ggplot defaults}
demo <- tribble(
    ~cut, ~freq,
    "Fair", 1610,
    "Good", 4906,
    "Very Good", 12082,
    "Premium", 13791,
    "Ideal", 21551
)
demo

ggplot(demo) +
    geom_bar(aes(x = cut, y = freq), stat = "identity")

# override default mapping from transformed variables to aesthetics
# e.g. display a bar chart of the PROPORTION of your total diamond dataset rather than a count
ggplot(diamonds) +
    geom_bar(aes(x = cut, y = stat(prop), group = 1))

# you can add statistical details to plots:
ggplot(diamonds) +
    stat_summary(
        mapping = aes(x = cut, y = depth),
        fun.min = min,
        fun.max = max,
        fun = median
    )
```

Adjusting the aesthetics (`aes()`) allows you to modify how your plot looks and what relationships it shows:

```{r Aesthetic adjustments}
# for bar plots, 'colour' changes the OUTLINE colour
ggplot(diamonds) +
    geom_bar(mapping = aes(x = cut, colour = cut))

# 'fill' changes the FILL colour
ggplot(diamonds) +
    geom_bar(mapping = aes(x = cut, fill = cut))

# default: position = "identity"
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity)) 

# alter transparency
ggplot(diamonds, aes(x = cut, fill = clarity)) +
    geom_bar(alpha = 1/5, position = "identity")

# colour bar outlines with no fill colour
ggplot(diamonds, aes(x = cut, colour = clarity)) +
    geom_bar(fill = NA, position = "identity")

# position = "fill" makes each set of stacked bars the same height
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "fill")

ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "dodge")

# position = "jitter" adds small amount of random noise to each point to avoid overplotting when points overlap...good for scatterplots, not barplots
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "jitter")
```
```{r penguins dataset}
str(penguins)

ggplot(penguins, aes(x = species, fill = species)) +
    geom_bar()

ggplot(penguins, aes(x = species, y = sex, fill = island)) +
    geom_col(position = "dodge")

ggplot(penguins, aes(x = species, y = flipper_length_mm, colour = species)) +
    geom_jitter() +
    theme(legend.position = "none")
```


# **Workshop 3: Reproducible Research in Marine Science**

```{r linking GitHub and RStudio}

library(usethis)

# in the console:
  # credentials::git_credential_ask()
  # usethis::git_sitrep()

# in the console:
  # usethis::use_git()

# after staging and committing files from the Git pane, in the console:
  # usethis::use_github()

# in RStudio terminal:
  # git remote add origin https://github.com/katiedyck/MB5370_Module01.git
  # git push -u origin master
# this created a master branch in the repository in GitHub, and the files `.gitignore` and `MB5370_Module01.git` are now visible

# in RStudio terminal:
  # git push
```


# **Workshop 4: Developers Toolbox & AI**

All 3 of the graphs below were created using Copilot, an AI coding assistant that can generate code based on natural language prompts. Comments in the code chunks indicate the prompts I used to generate the code for each graph.

## GitHub Copilot

```{r Copilot}

# Load tidyverse and palmerpenguins libraries
library(tidyverse)
library(palmerpenguins)

# Create a ggplot scatter plot using the penguins dataset. 
# Plot bill_length_mm on the x-axis and bill_depth_mm on the y-axis.
# Color points by species and use geom_point with size 3.
# Add a minimal theme and clean labels.
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Bill Length vs. Bill Depth of Palmer Penguins",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Species"
  )

# Create a ggplot boxplot using the penguins dataset.
# Compare flipper_length_mm across different species.
# Facet by island.
# Use custom color palettes.
ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ island) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(
    title = "Flipper Length of Palmer Penguins by Species and Island",
    x = "Species",
    y = "Flipper Length (mm)",
    fill = "Species"
  )  

# Create a ggplot scatterplot using the penguins dataset.
# Compare body_mass_g against flipper_length_mm.
# Add a linear regression trend line with geom_smooth(method = "lm") for each species.
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Body Mass vs. Flipper Length of Palmer Penguins",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  )  
```

## Google Gemini & `chattr`

```{r setting up & installing Google Gemini and chattr}

# in the console --> usethis::edit_r_environ()
  # in the console --> GEMINI_API_KEY="<my_custom_code>"
  # save the .Renviron file
# restart R (Command+Shift+0)


# Connecting `chattr` pkg to my model:
  # usethis::edit_r_profile()
    # options(.chattr_chat = ellmer::chat_google_gemini()) --> save the .Rprofile file
# restart R (Command+Shift+0)

# To interact with `chattr`, type into the console:
  # chattr::chattr_app()
# or use my keyboard shortcut (Ctrl + Command + C)
```

## `if()` / `else()`

```{r Base R: if and else}

# Simple Example: Checking Sea Surface Temperature (SST)
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}
```

## `if_else()` / `case_when()`

```{r dplyr: if_else(), case_when()}

# Sample coral data
coral_monitoring <- tibble(
  site = c("Site_A", "Site_B", "Site_C"),
  depth_m = c(12, 35, 8)
)

# Classify depth using if_else
coral_monitoring <- coral_monitoring %>% 
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef"))
coral_monitoring

  # mutate() creates a new column in the tibble based on the if_else() condition - if the reef is deeper than 30 meters it is classified as a Deep Reef, otherwise it is classified as a Shallow Reef

  # %>% is the pipe operator. It allows you to pass a series of expressions; it passes the result of the expression that's on the left of the pipe as the first argument for the right-hand side function.


coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))
coral_monitoring

# When you have more than two possibilities, stacking multiple if-else blocks becomes unreadable. case_when() provides a clean, sequential evaluation layout using formula tildes (~).
```

## Create a tibble with `chattr`

```{r Creating a tibble with chattr}

# Entered into a chattr chat: 
#   Write a script that creates a tibble named marine_stations that contains a column called salinity with values 35, 28, 32, and 12. Use the pipe operator (%>%) and mutate() combined with case_when() to create a new column called environment_type. Classify values below 15 as "Estuarine", values between 15 and 30 as "Brackish", and values above 30 as "Marine".

marine_stations <- tibble(salinity = c(35, 28, 32, 12)) %>%
  mutate(
    environment_type = case_when(
      salinity < 15 ~ "Estuarine",
      salinity >= 15 & salinity <= 30 ~ "Brackish",
      salinity > 30 ~ "Marine",
      TRUE ~ NA_character_ # Catch-all for any other cases, though not strictly needed here
    )
  )

marine_stations

```

## for loops

```{r Base R: for loops}

# Classic loop anatomy

for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}


# Loop across index sequences
transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}

number_turtles <- c(17, 39, 56, 44)
for (i in number_turtles) {
  print(paste("There were", i, "turtles on the beach last night."))
}
```

## Programming with the `purrr` package

```{r Tidyverse Alternative: Functional Programming (purrr)}

# Different map functions in the `purrr` package map a specific operation onto every item in a list or vector, automatically ensuring that your output matches the precise data structure you expect.

# Using a standard for-loop:
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])
}

# Using the purrr equivalent:
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)
mapped_results


# purrr pipelines can be combined inside data frames to split analyses across separate groups:
iris %>% 
  split(.$Species) %>% 
  map(~summary(.x))     # iterate a summary function over split lists of data


# From chattr prompt: "I have a list of three vectors containing fish count data. Write a for-loop to calculate the mean of each, and then show me the exact parallel way to do it using purrr's map_dbl function."

# Sample data
fish_counts_list <- list(
  river_a = c(10, 12, 8, 15, 11),
  lake_b = c(25, 22, 28, 20, 26, 24),
  stream_c = c(5, 7, 6, 9)
)

# Using a for-loop
mean_counts_for <- vector("numeric", length(fish_counts_list))
names(mean_counts_for) <- names(fish_counts_list)

for (i in seq_along(fish_counts_list)) {
  mean_counts_for[i] <- mean(fish_counts_list[[i]])
}

mean_counts_for

# Using purrr map function:
mean_counts_map <- map_dbl(fish_counts_list, mean)
mean_counts_map

```

## Writing custom functions

```{r Writing custom functions}

# Function Definition
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
  
}

# Utilizing your custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84)


# Custom function: convert temperature from degrees celcius to degrees fahrenheit
convert_temp_c_to_f <- function(C) {
  
  # conditional safety check
  if (C < -273.15) {
    stop("Cannot be below absolute zero!")
  }
  
  fahrenheit <- (C * 9/5) + 32
  #return(fahrenheit)
  print(paste(C, "\u00B0C =", round(fahrenheit, digits = 2), "\u00B0F."))
}

convert_temp_c_to_f(22)
convert_temp_c_to_f(50)
convert_temp_c_to_f(35)
convert_temp_c_to_f(-16)
```

## Advanced Scripting Challenges

All comments were used with `chattr` to generate my code.

```{r Scripting Challenge 1}
# -- Challenge 1: Nested Multi-Vector Iteration --

# In real-world field surveys, calculations often depend on matching values across multiple lists or matrices. Read the documentation for purrr::map2_dbl(). Given a vector of raw survey counts counts <- c(15, 24, 8, 42) and a vector of species-specific scaling factors factors <- c(1.2, 0.8, 2.5, 1.1), write a mapping statement that iterates across both vectors simultaneously to return a single vector of scaled abundance values. Do not use a manual index loop.

counts <- c(15, 24, 8, 42)
factors <- c(1.2, 0.8, 2.5, 1.1)

map2_dbl(counts, factors, .f = `*`)
```

```{r Scripting Challenge 2}
# -- Challenge 2: The Robust Data Wrangling Pipeline Function --

# Create a custom function named clean_and_classify_survey that handles complex data frame structures. The function must fulfill these criteria:
    # Accept a data frame and a column name threshold parameter as inputs.
    # Safely check if the specified column name exists within the passed data frame; if it does not, trigger a user-friendly error message using stop().
    # Filter out rows containing missing values (NA) within that specific column.
    # Use the pipe operator and mutate() along with a vectorised conditional (if_else() or case_when()) to append a status string variable classification based on your threshold parameter.
    # Return the cleaned, altered data frame to the global environment.

# Test your custom pipeline using a mock tibble that contains at least one missing value to confirm your code handles anomalies without crashing.

clean_and_classify_survey <- function(df, col_name, threshold) {
  # Safely check if the specified column name exists
  if (!col_name %in% names(df)) {
    stop(paste0("Error: Column '", col_name, "' not found in the data frame."))
  }

  # Convert col_name to a symbol for tidy evaluation
  col_sym <- sym(col_name)

  df %>%
    # Filter out rows containing missing values (NA) within that specific column
    filter(!is.na(!!col_sym)) %>%
    # Append a status string variable 'classification' based on the threshold
    mutate(classification = if_else(!!col_sym > threshold, "High", "Low"))
}

# Test your custom pipeline using a mock tibble
mock_survey_data <- tibble::tibble(
  plot_id = 1:7,
  species_count = c(15, 24, NA, 8, 42, 12, NA),
  habitat_type = c("Forest", "Grassland", "Forest", "Wetland", "Grassland", "Forest", "Wetland")
)

# Test 1: Successful run with valid column and NAs handled
cleaned_classified_data <- clean_and_classify_survey(mock_survey_data, "species_count", 20)
print(cleaned_classified_data)

# Test 2: Confirm your code handles anomalies without crashing (non-existent column)
# This will trigger the stop() error message
tryCatch({
  clean_and_classify_survey(mock_survey_data, "apple", 20)
}, error = function(e) {
  message(e$message)
})
```
