---
title: "Module 1: Programming Fundamentals"
author: "Katie Dyck"
date: "2026-05-20"
format:
  html:
      embed-resources: true
      code-fold: show
      code-copy: hover
      code-block-bg: true
      code-block-border-left: "grey25"
      code-overflow: wrap
      toc: true
      number-sections: true
      number-depth: 4
      toc-location: left
      toc-expand: 3
      toc-title: Contents
      syntax-highlighting: github
      theme: cosmo
      title-block-banner: true
      fig-height: 6
      fig-width: 10
      fig-cap-location: bottom
---

# Introduction

This qmd file includes all material covered from the "Introduction to Programming" workshops. The material covered includes:

- **Workshop 1: *A Basic Introduction to R***
  - Variables
  - Functions
  - Debugging code
  - Data types
  - Data structures
  - Data frames and tibbles
  - Installing and loading package libraries
  - Best coding practices
- **Workshop 2: *Data Visualization***
  - `ggplot2` package
  - Setting plot aesthetics (`aes()`) and geoms (`geom()`)
- **Workshop 3: *Reproducible Research in Marine Science***
  - Linking RStudio to GitHub
- **Workshop 4: *Developers Toolbox and AI***
  - Using GitHub's Copilot feature to help generate R code
  - Using Google Gemini and R `chattr` package to help generate R code
  - `if()` / `else()` / `if_else()` / `case_when()`
  - for loops
  - `purrr` package
  - How to write custom functions

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

# Housekeeping

Purge the global environment with the command `rm(list = ls())`.
```{r Clear session environment}
#| output: false

# Inventory every active object currently residing in session RAM
objects()

# Purge global environment
rm(list = ls())

# Confirm that global session memory is now vacant
objects()
```

# Load packages

Packages needed for this module are `tidyverse`, `palmerpenguins`, `purrr`, and `usethis`. 
```{r load packages}
library(tidyverse)
library(palmerpenguins)
library(purrr)
library(usethis)
```

# **Workshop 1: Introduction**

This workshop covered a basic introduction to coding in R.

```{r Workshop 1 Code, echo=TRUE}
#| output: false
#| code-fold: true

## ------ Getting Started ------ ####
# This section introduces us to R by using it as a simple calculator.

2 + 1
1:30
# 6 *
  # not complete - when you run this line, console asks you to unput a number after the asterisk

# 6 % 4  - doesn't work!
6 / 4



## ------ Variables and Assignment ------ ####
# This section focuses on understanding how data is stored in R and why that matters.

age <- 25
first_name <- 'Bill'

age
first_name

age + 1
age + age

# In R, you can’t use a number at the beginning of an object name and you can’t use some special symbols, including spaces.
# 01_age <- 25 # starts with a number
# !_age <- 25 # no special symbols
# age bob <- 25 # no spaces
Age <- 41 # cases matter

`age bob` <- 25 # no spaces, but with back ticks
# This is useful in some situations, however, it is better to use informative object names to make your code readable

# Run an algorithm:
25 + 15.1 + 20.25
x <- 30
25 + 15.1 + 20.25 + x



## ------ Functions ------ ####
# This section focuses on exploring built-in R functions

years_old <- 25.7
round(years_old)  # round up
floor(years_old)  # returns the largest integer not greater than elements of 'years_old'

years_old <- 25.765
round(years_old, digits = 2)

age <- 30
name <- "Katie"
paste(name, "is", age, "years old.", sep = " ")
paste(name, "is almost", age + 1, "years old.", sep = " ")



## ------ Debugging ------ ####
# Debugging is the process of finding and fixing errors in your code.

# Misconceptions
# variables don't work the same way as they do in spreadsheets

grade <- 55
total <- grade + 10
print(total)
grade <- 90
print(total)  # value of total in a spreadsheet would be 100, but here a variable holds the value it was assigned (65)
total <- grade + 10
print(total)

p <- 2
z <- 5
out <- p * z
print(out)

# debugging errors
# out <- p * a   # undefined variable 'a'
## > Error: object 'a' not found
out <- p * z

x <- 1
is.character(x)
is.integer(x)
is.numeric(x)
is.logical(x)

# my_quiz <- c("uno",
#              "dos",
#              "tres",
#              "cuatro"   # comma is missing!
#              "cinco")
# print(my_quis)   # spelling error
# str(my_quiz)
# len(my_quiz)   # len() not a function

my_quiz_fixed <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print(my_quiz_fixed)
str(my_quiz_fixed)
length(my_quiz_fixed)



## ------ Data types ------ ####

my_numeric_variable <- 4.2
class(my_numeric_variable)
typeof(my_numeric_variable)
attributes(my_numeric_variable)
length(my_numeric_variable)



## ------ Data structures ------ ####

# A vector is a collection of elements that are most commonly of the type character, numeric, integer, logical.
y <- c(1,2,3)
class(y)
typeof(y)
z <- c("Sarah", "Tracy", "John")
class(z)
typeof(z)

# Values inside a list can be several different types.
x <- list(1, "A", TRUE)
x
class(x)
typeof(x)
length(x)
x[[2]]   # use square brackets to retrieve individual elements in a list



## ------ Data frames and tibbles ------ ####
# 2D or "rectangular" data files

my_dataframe <- data.frame(
  no = c(1,2,3),
  name = c("Tracey", "John", "Pete"),
  logical = c(TRUE, FALSE, TRUE)
)
my_dataframe
str(my_dataframe)

my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)



## ------ Packages and libraries ------ ####
# Using software that other people have written

# install.packages("ggplot2", "tidyr", "earthtones", "dplyr")
# library(ggplot2)



## ------ Coding best practices ------ ####

# To take a .csv file and turn it into a plot of the abundance of a single species:
#   1. read the .csv file dataset into R - read_csv()
#   2. view the dataset - e.g. str('dataset_name_here'), head('dataset_name_here'), glimpse('dataset_name_here')
#   3. identify which columns/variables you want to use to create the plot, and which plot type (geom_*()) you want to create
#   4. create your plot with ggplot() + geom_*()
```


# **Workshop 2: Data Visualization**

This workshop covered different ways to visualize data in R. We used the dataset `mpg`, which contains information on the fuel economy of various car models from 1999 to 2008.

```{r mpg dataset, echo=TRUE}
#| code-fold: true

str(mpg)
# "displ" = engine size (litres)
# "hwy" = car's fuel efficiency (mpg)
```

## Create a `ggplot`

The `ggplot2` package library allows you to visualize different relationships that exist within one or multiple datasets. There are many different ways that plots can be customised when using this package.
```{r Basic ggplots, echo=TRUE}
#| code-fold: true

ggplot(mpg, aes(displ, hwy)) +
    geom_point()
```

Setting an `alpha` value changes the transparency of data points and lines:
```{r ggplot2 alpha, echo=TRUE}
#| code-fold: true

# change transparency by class (alpha)
ggplot(mpg, aes(displ, hwy, alpha = drv)) +
    geom_point()
```

You can set a specific `color` for one (or multiple) variables:
```{r ggplot2 color, echo=TRUE}
#| code-fold: true

# change colour by class
ggplot(mpg, aes(displ, hwy, color = drv)) +
    geom_point()
```

You change change the point `shape`:
```{r ggplot2 point shape, echo=TRUE}
#| code-fold: true

# change point shape by class
ggplot(mpg, aes(displ, hwy, shape = drv)) +
    geom_point()
```

You can change the point `size`:
```{r ggplot2 point size, echo=TRUE}
#| code-fold: true

# change point size by class
ggplot(mpg, aes(displ, hwy, size = drv)) +
    geom_point()
```

You can combine settings to change the appearance of variables and tell a story:
```{r ggplot2 shape size and color}
#| code-fold: true

# change shape, size, AND colour by class
ggplot(mpg, aes(displ, hwy, color = drv, shape = drv, size = drv)) +
    geom_point()

ggplot(mpg, aes(displ, hwy, colour = displ < 5)) +
    geom_point()
```

`facet_wrap()` and `facet_grid()` allow you to visualize the data in ways that make relationships between variables (particularly those that are discrete or factored) clearer.

```{r facet_wrap() and facet_grid(), echo=TRUE}
#| code-fold: true

# ggplot(mpg, aes(displ, hwy, colour = drv)) +
#     geom_jitter() +
#     facet_wrap(~ drv) +
#     theme(legend.position = "none",
#           plot.title = element_text(face = "bold")) +
#     labs(x = "Engine displacement (L)",
#          y = "Highway (mpg)")

# ggplot(mpg) +
#     geom_point(aes(displ, hwy)) +
#     facet_wrap(~ class, nrow = 2)

# facet_grid() is used with more than one variable
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(drv ~ cyl)

# use a . if you don't want to facet in the rows or columns dimension
# ggplot(mpg) +
#     geom_point(aes(displ, hwy)) +
#     facet_grid(. ~ cyl)
ggplot(mpg) +
    geom_point(aes(displ, hwy)) +
    facet_grid(drv ~ .)
```

`geom_smooth()` makes it easier to visualize patterns in the data, particularly when there are many data points present.

-   `loess` (Locally Estimated Scatterplot Smoothing) line - fits a smooth, localised regression line that bends to show non-linear patterns.
```{r, echo=TRUE}
#| code-fold: true

ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth(se = FALSE) +   # method = "loess" (default)
    labs(caption = "Smoothing method = `loess`")
```

-   `lm` (Linear Model) line - fits a single, straight linear regression line across the entire dataset.
```{r, echo=TRUE}
#| code-fold: true

ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(caption = "Smoothing method = `lm`")
```

`geom_smooth()` can accept a variable for the `linetype` argument, which allows you to visualize how the data changes for different values of a specific variable. The graph below shows how highway miles per gallon changes with engine displacement (in litres) between different types of drive train (4-wheel drive, front-wheel drive, and rear-wheel drive).

```{r, echo=TRUE}
#| code-fold: true

ggplot(mpg) +
    geom_smooth(aes(displ, hwy, linetype = drv))
```

You can create plots with multiple geoms to show relationships between different variables in the data:

```{r Using multiple geoms, echo=TRUE}
#| code-fold: true

ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = drv)) +
    geom_smooth(aes(displ, hwy, colour = drv, linetype = drv))
```

You can use mapping in specific layers to enable the display of different aesthetics in different layers. In the second plot below, the aesthetics of the `geom_smooth()` layer have been filtered to only include subcompact cars.

```{r Advanced plots, echo=TRUE}
#| code-fold: true

ggplot(mpg, aes(displ, hwy)) +
    geom_smooth() +
    geom_point(aes(colour = class))

ggplot(mpg, aes(displ, hwy)) +
    geom_point(aes(colour = class)) +
    geom_smooth(data = filter(mpg, class == "subcompact")) +
    labs(caption = "Data has been filtered for class 'subcompact'")
```

## Transformations and Stats

The `diamonds` dataset (from the ggplot2 package) contains information on over 50,000 cut diamonds.

```{r diamonds dataset, echo=TRUE}
#| code-fold: true

str(diamonds)
```

```{r diamonds plot 1, echo=TRUE}
ggplot(diamonds) +
  geom_bar(aes(x = cut))
```
"Count" is not a variable in the `diamonds` dataset. Rather, the graph includes raw values of the dataset. Many graph types calculate new values to plot:

- **bar charts**, **histograms**, and **frequency polygrams** "bin" data and then plot bin counts (the number of data points that fall in each bin).
- **smoothers** fit a model to the data and then plot predictions from the model.
- **boxplots** compute a summary of the distribution and then display a specially formatted box.

`geoms` and `stats` can generally be used interchangeably.

```{r diamonds plot 2, echo=TRUE}
ggplot(diamonds) +
    stat_count(mapping = aes(x = cut))
```
The tibble below contains the frequency of each diamond `cut` that appears in the dataset. We can use this table with the `stat` argument to override the default mapping.

```{r Overriding ggplot defaults, echo=TRUE}
#| code-fold: true

demo <- tribble(
    ~cut, ~freq,
    "Fair", 1610,
    "Good", 4906,
    "Very Good", 12082,
    "Premium", 13791,
    "Ideal", 21551
)
demo

ggplot(demo) +
    geom_bar(aes(x = cut, y = freq), stat = "identity")

# override default mapping from transformed variables to aesthetics
# e.g. display a bar chart of the PROPORTION of your total diamond dataset rather than a count
ggplot(diamonds) +
    geom_bar(aes(x = cut, y = stat(prop), group = 1))
```

You can add statistical details to plots (e.g. min, max, median):
```{r, echo=TRUE}
#| code-fold: true

ggplot(diamonds) +
    stat_summary(
        mapping = aes(x = cut, y = depth),
        fun.min = min,
        fun.max = max,
        fun = median
    )
```

Adjusting the aesthetics (`aes()`) allows you to modify how your plot looks.

```{r Aesthetic adjustments, echo=TRUE}
# for bar plots, 'colour' changes the OUTLINE colour
ggplot(diamonds) +
    geom_bar(mapping = aes(x = cut, colour = cut))

# 'fill' changes the FILL colour
ggplot(diamonds) +
    geom_bar(mapping = aes(x = cut, fill = cut))

# default: position = "identity"
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity)) 

# alter transparency
ggplot(diamonds, aes(x = cut, fill = clarity)) +
    geom_bar(alpha = 1/5, position = "identity")

# colour bar outlines with no fill colour
ggplot(diamonds, aes(x = cut, colour = clarity)) +
    geom_bar(fill = NA, position = "identity")

# position = "fill" makes each set of stacked bars the same height
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "fill")

ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "dodge")

# position = "jitter" adds small amount of random noise to each point to avoid overplotting when points overlap...good for scatterplots, not barplots
ggplot(diamonds) +
    geom_bar(aes(x = cut, fill = clarity), position = "jitter")
```


# **Workshop 3: Reproducible Research in Marine Science**

::: {.callout-note icon=FALSE}
The R package library `usethis` **must** be installed and loaded in order to link your GitHub account to RStudio.
:::

```{r linking GitHub and RStudio, echo=TRUE}
#| code-fold: true

# in the console:
  # credentials::git_credential_ask()
  # usethis::git_sitrep()

# in the console:
  # usethis::use_git()

# after staging and committing files from the Git pane, in the console:
  # usethis::use_github()

# in RStudio terminal:
  # git remote add origin https://github.com/katiedyck/MB5370_Module01.git
  # git push -u origin master
# this created a master branch in the repository in GitHub, and the files `.gitignore` and `MB5370_Module01.git` are now visible

# in RStudio terminal:
  # git push
```


# **Workshop 4: Developers Toolbox & AI**

## GitHub Copilot

All 3 of the graphs below were created using Copilot, an AI coding assistant that can generate code based on natural language prompts. Comments in the code chunks indicate the prompts used to generate the code for each graph.

```{r Copilot, echo=TRUE}
#| code-fold: true

# Load tidyverse and palmerpenguins libraries
library(tidyverse)
library(palmerpenguins)

# Create a ggplot scatter plot using the penguins dataset. 
# Plot bill_length_mm on the x-axis and bill_depth_mm on the y-axis.
# Color points by species and use geom_point with size 3.
# Add a minimal theme and clean labels.
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Bill Length vs. Bill Depth of Palmer Penguins",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Species"
  )

# Create a ggplot boxplot using the penguins dataset.
# Compare flipper_length_mm across different species.
# Facet by island.
# Use custom color palettes.
ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ island) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(
    title = "Flipper Length of Palmer Penguins by Species and Island",
    x = "Species",
    y = "Flipper Length (mm)",
    fill = "Species"
  )  

# Create a ggplot scatterplot using the penguins dataset.
# Compare body_mass_g against flipper_length_mm.
# Add a linear regression trend line with geom_smooth(method = "lm") for each species.
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Body Mass vs. Flipper Length of Palmer Penguins",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  )  
```

## Google Gemini & `chattr`

The code below was run in the R Console to install Google Gemini and the `chattr` library before connecting them to RStudio.
```{r setting up & installing Google Gemini and chattr, echo=TRUE}
#| code-fold: true

# in the console --> usethis::edit_r_environ()
  # in the console --> GEMINI_API_KEY="<my_custom_code>"
  # save the .Renviron file
# restart R (Command+Shift+0)


# Connecting `chattr` pkg to my model:
  # usethis::edit_r_profile()
    # options(.chattr_chat = ellmer::chat_google_gemini()) --> save the .Rprofile file
# restart R (Command+Shift+0)

# To interact with `chattr`, type into the console:
  # chattr::chattr_app()
# or use my keyboard shortcut (Ctrl + Command + C)
```

## `if()` / `else()`

```{r Base R: if and else, echo=TRUE}
# Simple Example: Checking Sea Surface Temperature (SST)
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}
```

## `if_else()` / `case_when()`

```{r dplyr if_else(), echo=TRUE}
# Sample coral data
coral_monitoring <- tibble(
  site = c("Site_A", "Site_B", "Site_C"),
  depth_m = c(12, 35, 8)
)

# Classify depth using if_else
coral_monitoring <- coral_monitoring %>% 
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef"))
coral_monitoring
```

-   `mutate()` creates a new column in the tibble based on the if_else condition - if the reef is deeper than 30 meters it is classified as a Deep Reef, otherwise it is classified as a Shallow Reef.

-   `%>%` is the **pipe operator**. It allows you to pass a series of expressions; it passes the result of the expression that's on the left of the pipe as the first argument for the right-hand side function.


```{r dplyr case_when(), echo=TRUE}
coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))
coral_monitoring
```

When you have more than two possibilities, stacking multiple if-else blocks becomes unreadable. case_when() provides a clean, sequential evaluation layout using formula tildes (`~`).

## Create a tibble with `chattr`

The code below was creating by entering the comment into `chattr`.
```{r Creating a tibble with chattr, echo=TRUE}
# Write a script that creates a tibble named marine_stations that contains a column called salinity with values 35, 28, 32, and 12. Use the pipe operator (%>%) and mutate() combined with case_when() to create a new column called environment_type. Classify values below 15 as "Estuarine", values between 15 and 30 as "Brackish", and values above 30 as "Marine".

marine_stations <- tibble(salinity = c(35, 28, 32, 12)) %>%
  mutate(
    environment_type = case_when(
      salinity < 15 ~ "Estuarine",
      salinity >= 15 & salinity <= 30 ~ "Brackish",
      salinity > 30 ~ "Marine",
      TRUE ~ NA_character_ # Catch-all for any other cases, though not strictly needed here
    )
  )

marine_stations
```

## for loops

```{r Base R: for loops, echo=TRUE}
# Classic loop anatomy
for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}


# Loop across index sequences
transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}

number_turtles <- c(17, 39, 56, 44)
for (i in number_turtles) {
  print(paste("There were", i, "turtles on the beach last night."))
}
```

## Programming with the `purrr` package

Different map functions in the `purrr` package map a specific operation onto every item in a list or vector, automatically ensuring that your output matches the precise data structure you expect.

```{r Tidyverse Alternative: Functional Programming (purrr), echo=TRUE}
# Using a standard for-loop:
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas))  # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])
}
results

# Using the purrr equivalent:
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)
mapped_results


# purrr pipelines can be combined inside data frames to split analyses across separate groups:
iris %>% 
  split(.$Species) %>% 
  map(~summary(.x))     # iterate a summary function over split lists of data


# From chattr prompt: "I have a list of three vectors containing fish count data. Write a for-loop to calculate the mean of each, and then show me the exact parallel way to do it using purrr's map_dbl function."

# Sample data
fish_counts_list <- list(
  river_a = c(10, 12, 8, 15, 11),
  lake_b = c(25, 22, 28, 20, 26, 24),
  stream_c = c(5, 7, 6, 9)
)

# Using a for-loop
mean_counts_for <- vector("numeric", length(fish_counts_list))
names(mean_counts_for) <- names(fish_counts_list)

for (i in seq_along(fish_counts_list)) {
  mean_counts_for[i] <- mean(fish_counts_list[[i]])
}

mean_counts_for

# Using purrr map function:
mean_counts_map <- map_dbl(fish_counts_list, mean)
mean_counts_map

```

## Writing custom functions

```{r Writing custom functions, echo=TRUE}
# Function Definition
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
  
}

# Utilizing your custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84)


# Custom function: convert temperature from degrees celcius to degrees fahrenheit
convert_temp_c_to_f <- function(C) {
  
  # conditional safety check
  if (C < -273.15) {
    stop("Cannot be below absolute zero!")
  }
  
  fahrenheit <- (C * 9/5) + 32
  #return(fahrenheit)
  print(paste(C, "\u00B0C =", round(fahrenheit, digits = 2), "\u00B0F."))
}

convert_temp_c_to_f(22)
convert_temp_c_to_f(50)
convert_temp_c_to_f(35)
convert_temp_c_to_f(-16)
```

## Advanced Scripting Challenges

The paragraphs in the comments below were entered into `chattr` to generate my code.

### Challenge 1: Nested Multi-Vector Iteration

```{r Scripting Challenge 1, echo=TRUE}
# Given a vector of raw survey counts, counts <- c(15, 24, 8, 42), and a vector of species-specific scaling factors, factors <- c(1.2, 0.8, 2.5, 1.1), write a mapping statement that iterates across both vectors simultaneously to return a single vector of scaled abundance values. Do not use a manual index loop.

counts <- c(15, 24, 8, 42)
factors <- c(1.2, 0.8, 2.5, 1.1)

map2_dbl(counts, factors, .f = `*`)
```

### Challenge 2: The Robust Data Wrangling Pipeline Function

```{r Scripting Challenge 2, echo=TRUE}
# Create a custom function, "clean_and_classify_survey", that can handle complex data frame structures. The function must be able to: accept a data frame (df) and a column name threshold parameter as inputs; safely check if the specified column name exists within the passed data frame, if not it triggers a user-friendly error message using stop(); filter out rows containing missing values (NA) within the specific column; use the pipe operator and mutate() along with a vectorised conditional (if_else() or case_when()) to append a status string variable classification based on the threshold parameter; return the cleaned, altered data frame to the global environment. Test the custom pipeline using a mock tibble that contains at least one missing value to confirm the code handles anomalies without crashing.

clean_and_classify_survey <- function(df, column_name, threshold) {
  # Safely check if the specified column name exists
  if (!column_name %in% names(df)) {
    stop(paste0("Error: Column '", column_name, "' not found in the data frame."))
  }

  # Convert column_name to a symbol for tidy evaluation
  col_sym <- sym(column_name)

  df_cleaned <- df %>%
    # Filter out rows containing missing values (NA) within the specific column
    filter(!is.na(!!col_sym)) %>%
    # Append a status string variable classification based on the threshold
    mutate(status = if_else(!!col_sym >= threshold, "High Performer", "Low Performer"))

  return(df_cleaned)
}

# Create a mock tibble for testing
mock_survey_data <- tibble(
  respondent_id = 101:105,
  satisfaction_score = c(85, NA, 92, 78, 60),
  feedback_category = c("Positive", "Neutral", "Positive", "Negative", "Neutral"),
  completion_date = as.Date(c("2023-10-26", "2023-10-27", "2023-10-28", "2023-10-29", "2023-10-30"))
)
print(mock_survey_data)

# Test 1: Happy path with existing column and NAs
cleaned_data_1 <- clean_and_classify_survey(
  df = mock_survey_data,
  column_name = "satisfaction_score",
  threshold = 80
)
print(cleaned_data_1)

# Test 2: Verify error handling for a non-existent column
# This will stop execution and print the custom error message
    # cleaned_data_2 <- clean_and_classify_survey(
    #   df = mock_survey_data,
    #   column_name = "non_existent_column",
    #   threshold = 80
    # )
```
